---
title: 'Exploring ElderNet Datasets'
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, fig.height = 7, fig.width = 9, message = FALSE, warning = FALSE)
options(scipen = 100)

library(tidyverse)
library(lubridate)
library(purrr)
library(skimr)
library(here)
library(tidyr)
library(stringr)
library(ggsci)

# some custom functions
round_numerics <- 
  function(data, digits = 2){
    data %>%
      mutate(across(where(is.numeric), ~ round(.x, digits)))
  }

add_table <- 
  function(data){
    data %>%
      round_numerics() %>%
      reactable::reactable(., fullWidth = F, resizable = T, filterable = T, 
                           highlight = T, defaultPageSize = 10, wrap = FALSE,
                           showSortIcon = T, striped = T, compact = T)
  }

bg_theme <- 
  function(base_size = 11,
           strip_text_size = 12,
           strip_text_margin = 10,
           subtitle_size = 13,
           subtitle_margin = 10,
           plot_title_size = 14,
           plot_title_margin = 10,
           font = "RobotoMono-Regular",
           ...) {
    
    ret <-
      ggplot2::theme_gray(base_family = font,
                          base_size = base_size, ...,) +
      theme(
        panel.background = element_rect(fill = "#e9e9ea"),
        plot.background = element_rect(fill = "#f3f3f3"),
        legend.background = element_rect(fill = "#f3f3f3"),
        panel.grid = element_line(color = "#ffffff"),
        panel.grid.major = element_line(color = "#ffffff")
        )
    
    ret$strip.text <-
      ggplot2::element_text(
        # hjust = 0,
        vjust = -.8,
        size = strip_text_size,
        margin = ggplot2::margin(b = strip_text_margin),
        family = font
      )
    
    ret$plot.subtitle <-
      ggplot2::element_text(
        hjust = 0,
        size = subtitle_size,
        margin = ggplot2::margin(b = subtitle_margin),
        family = font
      )
    
    ret$plot.title <-
      ggplot2::element_text(
        hjust = 0,
        size = plot_title_size,
        margin = ggplot2::margin(b = plot_title_margin),
        family = font
      )
    
    ret
  }

bg_red <- "#E64B35"
bg_green <- "#00a087"
bg_blue <- "#4DBBD5"

#load data
path <-
  list.files(path = here("data"), pattern = ".csv")

file_names <-
  c("care_mgmt", "client_info", "donations", "pantry", "volunteer")

path %>% 
  map( ~ read_csv(here("data", .))) %>% 
  set_names(file_names) %>%
  list2env(., envir = .GlobalEnv)
  
```

### Client Info

main takeaways:

 * There are 284 Client IDs are duplicated in the Client_info dataset. Example: anon_ID = `625` has 3 values for `age_group`, `T`, `D` and `missing`
 * There is lots of missing demographic data for poverty status, minority status and age group

```{r}
# client_info %>% 
#   add_table()
```

```{r}

duplicated <- 
  client_info %>%
  group_by(anon_ID) %>%
  tally(sort = TRUE) %>%
  filter(n > 1) 

duplicated %>%
  add_table()

```

```{r}
client_info %>%
  group_by(poverty, minority, age_group) %>% 
  tally() %>%
  mutate(label = paste('poverty:', poverty),
         age_group = ifelse(is.na(age_group), 'missing', age_group),
         minority = ifelse(is.na(minority), 'missing', minority)) %>%
  ggplot(., aes(x = age_group, y = n, fill = minority)) + 
  geom_col() + 
  facet_wrap(vars(label)) + 
  bg_theme() + 
  scale_fill_npg() + 
  labs(title =" ElderNet Clients by Age Group, Poverty Level and Mintory Status")

```

### Care Management

```{r}
care_mgmt %>% add_table()
```

Care management encounters per client is right-skewed. The max is 371 encounters, min is 1 and the median number of care management encounters per client is 6.

```{r}

# care_mgmt %>%
#   group_by(anon_ID) %>%
#   tally(sort = TRUE)  %>%
#   add_table()
```

```{r}
quantiles <- 
  quantile(care_mgmt %>% group_by(anon_ID) %>% tally() %>% pull(n), probs = seq(0, 1, .10))

quantiles

```

```{r}

care_mgmt %>%
  group_by(anon_ID) %>%
  tally() %>% 
  ggplot(aes(x = n)) +
  geom_histogram(fill = bg_green, alpha = .75, bins = 50)+ 
  bg_theme() + 
  labs(title = "distribution of care management encounters") 

care_mgmt %>%
  mutate(month = floor_date(assistance_date, 'month')) %>%
  group_by(anon_ID, month) %>%
  tally() %>% 
  ggplot(aes(x = month, y = n, group = month)) +
  geom_jitter(color = 'gray', alpha = .55) + 
  geom_boxplot(fill = bg_green, alpha = .75, outlier.alpha = 0) + 
  labs(title = "distribution of care management encounters per month") + 
  bg_theme() + 
  scale_fill_npg() 

```

There is a decrease present in both `Community` and `Home Visit` encounters starting in March 2020. It also appears remote contacts with clients was shifted to more remote contacts without clients once the pandemic hit.

```{r}


care_mgmt %>%
  mutate(month = floor_date(assistance_date, 'month')) %>%
  group_by(assistance_category, month) %>%
  tally() %>% 
  padr::pad(group = "assistance_category") %>%
  padr::fill_by_value(n, 0) %>%
  ggplot(aes(x = month, y = n, color = assistance_category)) +
  geom_point() + 
  geom_line()+ 
  bg_theme() + 
  theme(legend.position = 'bottom') + 
  labs(color = NULL) +
  scale_color_npg()  + 
  labs(title = "monthly total care management encounters by type of service provided")

```

The average time spent per encounter varies from month to month.
```{r}

care_mgmt %>%
  mutate(month = floor_date(assistance_date, 'month')) %>%
  group_by(month, assistance_category) %>%
  summarise(mean_time_spent = mean(amount)) %>%
  padr::pad(group = "assistance_category") %>%
  padr::fill_by_value(mean_time_spent, 0) %>%
  ggplot(aes(x = month, y = mean_time_spent, color = assistance_category)) +
  geom_point(show.legend = FALSE) + 
  geom_line(show.legend = FALSE) + 
  facet_wrap(vars(assistance_category), scales = 'free') + 
  bg_theme() + 
  scale_color_npg()  + 
  labs(title = "mean time spent per encounter type per month")

```

### Pantry

```{r}
# pantry %>% add_table()
```

There is some repeating cycles present, likely due to holiday baskes (see next chart).
```{r}

pantry %>%
  mutate(assistance_date = mdy_hm(assistance_date),
         month = floor_date(assistance_date, 'month')) %>%
  group_by(month) %>%
  summarise(total = sum(amount)) %>%
  padr::pad() %>%
  padr::fill_by_value(n, 0) %>%
  ggplot(aes(x = month, y = total)) +
  geom_line(color = bg_green) + 
  geom_point(color = bg_green) + 
  bg_theme() +
  scale_color_npg() +
  labs(title = "total monthly pantry poundage")

```


```{r}
pantry %>%
  mutate(assistance_date = mdy_hm(assistance_date),
         month = floor_date(assistance_date, 'month')) %>%
  group_by(month, assistance_category) %>%
  tally() %>% 
  padr::pad(group = "assistance_category") %>%
  padr::fill_by_value(n, 0) %>%
  ggplot(aes(x = month, y = n, color = assistance_category)) +
  geom_line() + 
  geom_point()+ 
  bg_theme() +
  scale_color_npg() +
  labs(title = "Pantry poundage by category")

```

### Volunteers

```{r}
volunteer %>% add_table()
```

```{r}
volunteer %>%
  mutate(appt_date = mdy(appt_date),
         month = floor_date(appt_date, 'month')) %>%
  group_by(month, category) %>% 
  summarise(mean_duration = mean(appt_duration)) %>%
  padr::pad(group = "category") %>%
  padr::fill_by_value(mean_duration, 0) %>%
  ggplot(., aes(x = month, y = mean_duration, color = category)) +
  geom_line(show.legend = FALSE) + 
  geom_point(show.legend = FALSE) + 
  facet_wrap(vars(category), scales = 'free') +
  labs(title = "monthly volunteer duration by category")+ 
  bg_theme() +
  scale_color_npg() 
```

## Ideas

client story: 
* filterable UI w/ summary of:
  * number of recent interactions
  * time spent
  * pantry items
  * services utilized
* all absolute and relative to other active clients
* can track over time...

utilization metric:
* time in month  +  food + care coordination...
* track over time to see if more care is needed if metric starts to decline?

Frame ldernet benefits in terms of the alternative

 * ElderNet is the dignified alternative - providing access to care and food keeps people in their homes longer
 * ElderNet is the preferred alternative - providing access to care and food allows clients to remain in the home longer
 * ElderNet is the less expensive alternative - providing access to care keeps up clients health and can help prevent costly care due to delays in care
 * ElderNet is the alternative

```{r}

client <- 
  "210"

care_mgmt_story <- 
  care_mgmt %>%
  # filter(anon_ID == client) %>%
  mutate(month = as.Date(floor_date(assistance_date, 'month'))) %>%
  group_by(anon_ID, month) %>%
  summarise(total_care_mgmt_encounters = n(),
            total_care_mgmt_mins = sum(amount, na.rm = T),
            mean_care_mgmt_mins = mean(amount, na.rm = T))

pantry_story <- 
  pantry %>%
  # filter(anon_ID == client) %>%
  mutate(month = as.Date(floor_date(mdy_hm(assistance_date), 'month'))) %>%
  group_by(anon_ID, month) %>%
  summarise(total_pantry_encounters = n(),
            total_pantry_pounds = sum(amount, na.rm = T),
            mean_pantry_pounds = mean(amount, na.rm = T))

volunteer_story <- 
  volunteer %>%
  # filter(anon_ID == client) %>%
  mutate(month = as.Date(floor_date(mdy(appt_date), 'month'))) %>%
  group_by(anon_ID, month) %>%
  summarise(total_volunteer_encounters = n(),
            total_volunteer_mins = sum(appt_duration, na.rm = T),
            mean_volunteer_mins = mean(appt_duration, na.rm = T))

get_utilziation <- 
  function(client){
    
    client_info %>%
      filter(anon_ID == {{client}}) %>%
      left_join(., care_mgmt_story) %>%
      left_join(., pantry_story) %>%
      left_join(., volunteer_story) %>% 
      group_by(month) %>%
      summarise(encounters = sum(total_care_mgmt_encounters, total_pantry_encounters, total_volunteer_encounters, na.rm = T),
                mins = sum(total_care_mgmt_mins, mean_volunteer_mins, na.rm = T),
                food = total_pantry_pounds) %>% 
      mutate(monthly_total = sum(encounters, mins, food, na.rm = T))

  }

# total <- 
#   map_dfr(unique(client_info$anon_ID), ~get_utilziation(client = .x))

client_totals <-
  client_info %>%
  select(anon_ID) %>%
  left_join(., care_mgmt_story, by = c("anon_ID")) %>%
  left_join(., pantry_story, by = c("anon_ID", "month")) %>%
  left_join(., volunteer_story, by = c("anon_ID", "month")) 

client_totals %>%
  select(anon_ID, month, starts_with("mean")) %>%
  pivot_longer(cols = 3:5) %>%
  group_by(month, name) %>%
  summarise(total = sum(value, na.rm = T)) %>%
  ggplot(aes(x = month, y = total, fill = name)) + 
  geom_area()+ 
  bg_theme() +
  scale_fill_npg() 

client_totals %>%
  select(anon_ID, month, ends_with("encounters")) %>%
  select(-starts_with("mean")) %>%
  pivot_longer(cols = 3:5) %>%
  group_by(month, name) %>%
  summarise(total = sum(value, na.rm = T)) %>%
  ggplot(aes(x = month, y = total, fill = name)) +
  geom_area()+ 
  bg_theme() +
  scale_fill_npg() +
  labs(title = "Encounters")

client_totals %>%
  select(anon_ID, month, ends_with("mins")) %>%
  select(-starts_with("mean")) %>%
  pivot_longer(cols = 3:4) %>%
  group_by(month, name) %>%
  summarise(total = sum(value, na.rm = T)) %>%
  ggplot(aes(x = month, y = total, fill = name)) + 
  geom_area()+ 
  bg_theme() +
  scale_fill_npg() +
  labs(title = "Minutes")

client_totals %>%
  select(anon_ID, month, ends_with("pounds")) %>%
  select(-starts_with("mean")) %>%
  group_by(month) %>%
  summarise(total = sum(total_pantry_pounds, na.rm = T)) %>%
  ggplot(aes(x = month, y = total)) + 
  geom_area(fill = bg_green, alpha = .65)+ 
  bg_theme() +
  labs(title = "total pantry pounds")

```



