---
title: 'ElderNet Utilziation Working Doc'
output: github_document
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = T, fig.height = 7, fig.width = 9, message = FALSE, warning = FALSE)
options(scipen = 100)

library(tidyverse)
library(lubridate)
library(purrr)
library(skimr)
library(here)
library(tidyr)
library(stringr)
library(ggsci)
library(infer)
library(broom)
library(tidymodels)

```

```{r}
#load data
path <-
  list.files(path = here("data"), pattern = ".csv")

file_names <-
  c("care_mgmt", "client_info", "donations", "pantry", "volunteer")

path %>% 
  map( ~ read_csv(here("data", .))) %>% 
  set_names(file_names) %>%
  list2env(., envir = .GlobalEnv)

```

## prep the data
Not every client has an interaction with each service each month. For example, a client may use volunteer services this month, then in 2 months later visit the pantry leaving a 1 month gap in their utilization. To get a tidy dataset where we have 1 row per client per month, I created a 'master calendar' for each client, starting from the first date in the data set and ending at the last date. Later on I adjust this for the date of each client's first interaction.

```{r}

global_min <-
  min(
    c(
      min(as.Date(care_mgmt$assistance_date)),
      min(as.Date(mdy_hm(pantry$assistance_date))),
      min(as.Date(mdy(volunteer$appt_date)))
    )
  )

global_min

```

```{r}

global_max <- 
  max(
    c(
      max(as.Date(care_mgmt$assistance_date)),
      max(as.Date(mdy_hm(pantry$assistance_date))),
      max(as.Date(mdy(volunteer$appt_date)))
    )
  )

global_max

```

"Padding" the client data set results in each client having the same number of observations - each client has the same number of rows now representing the number of days between the min and max dates:

```{r}
client_info_adj <- 
  client_info %>%
  select(anon_ID) %>%
  mutate(date = as.Date("2019-01-01")) %>%
  group_by(anon_ID) %>%
  padr::pad(group = "anon_ID", interval = "day", start_val = as.Date(global_min), end_val = as.Date(global_max))

# each client now has 1 entry per day
client_info_adj %>% 
  group_by(anon_ID) %>% 
  tally(name = "days")

```

Rolling up to the month, now each client has 1 observation per month. Prepping the data this way will allow us to join the care mgmt, pantry and volunteer data to the client data for each client for each month, which will preserve instances where a client had 0 interactions in a given month.

```{r}

client_info <- 
  client_info %>%
  inner_join(., client_info_adj, "anon_ID") %>%
  mutate(month = floor_date(date, 'month')) %>%
  select(-date) %>%
  ungroup() %>%
  distinct()

# rolled up into months - each client has 1 entry per 33 months
client_info %>% 
  group_by(anon_ID) %>% 
  tally(name = 'months')
```

Example output:

```{r}
client_info %>% 
  filter(anon_ID == "210")

```

## get ElderNet encounters

this section counts client encounters for each month for each service offered. We can join all of these with the client data to try to summarize utilization.

#### care management
```{r}

care_mgmt_util <-
  care_mgmt %>%
  mutate(month = as.Date(floor_date(assistance_date, 'month'))) %>%
  group_by(anon_ID, month) %>%
  summarise(total_care_mgmt_encounters = n(),
            total_care_mgmt_mins = sum(amount, na.rm = T),
            mean_care_mgmt_mins = mean(amount, na.rm = T))

#example output for client 210
care_mgmt_util %>% 
  filter(anon_ID == "210")

```

#### pantry
```{r}

pantry_util <- 
  pantry %>%
  mutate(month = as.Date(floor_date(mdy_hm(assistance_date), 'month'))) %>%
  group_by(anon_ID, month) %>%
  summarise(total_pantry_encounters = n(),
            total_pantry_pounds = sum(amount, na.rm = T),
            mean_pantry_pounds = mean(amount, na.rm = T))

#example output for client 210
pantry_util %>% 
  filter(anon_ID == "210")

```

#### volunteer services
```{r}

volunteer_util <- 
  volunteer %>%
  mutate(month = as.Date(floor_date(mdy(appt_date), 'month'))) %>% 
  group_by(anon_ID, month) %>%
  summarise(total_volunteer_encounters = n(),
            total_volunteer_mins = sum(appt_duration, na.rm = T),
            mean_volunteer_mins = mean(appt_duration, na.rm = T))

#example output for client 210
volunteer_util %>% 
  filter(anon_ID == "210")

```

### get utilziation

Here all datasets are joined together and then services used for each client for each month are calculated. There are columns for the total encounters as well as binary indicator columns for if each type of service was used or not (regardless of how much of that service was used).

```{r}
monthly_util <-
  client_info %>%
  left_join(., care_mgmt_util, by = c("anon_ID", "month")) %>%
  left_join(., pantry_util, by = c("anon_ID", "month")) %>% 
  left_join(., volunteer_util, by = c("anon_ID", "month")) %>% 
  ungroup() %>%
  as_tibble() %>% 
  rowwise() %>% 
  mutate(across(where(is.numeric), ~(ifelse(is.na(.x), 0, .)))) %>%
  mutate(used_care_mgmt = ifelse(total_care_mgmt_encounters >= 1, 1, 0),
         used_pantry = ifelse(total_pantry_encounters >= 1, 1, 0),
         used_volunteer = ifelse(total_volunteer_encounters >=1, 1, 0)) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(num_svcs_used = sum(used_care_mgmt, used_pantry, used_volunteer, na.rm = T)) %>%
  as_tibble()
```

Finally, we need to adjust the utilization dataset to take into account each client's first month. We can remove all rows before each client's first date:

```{r}

global_min_client_dates <- 
  care_mgmt %>%
  mutate(assistance_date = as.Date(assistance_date)) %>%
  select(anon_ID, care_mgmt_assistance_date = assistance_date) %>%
  left_join(., pantry %>%
              mutate(assistance_date = as.Date(assistance_date, format = '%m/%d/%Y')) %>% 
              select(anon_ID, pantry_assistance_date = assistance_date)) %>%
  left_join(., volunteer %>% 
              mutate(rider_first_ride_date = as.Date(rider_first_ride_date)) %>%
              select(anon_ID, rider_first_ride_date)) %>%
  distinct() %>%
  group_by(anon_ID) %>%
  summarize(min_care_mgmt_assistance_date = min(care_mgmt_assistance_date, na.rm = TRUE),
            min_pantry_assistance_date = min(pantry_assistance_date, na.rm = TRUE),
            min_rider_first_ride_date = min(rider_first_ride_date, na.rm = TRUE))

global_min_client_dates <- 
  tibble(
    anon_ID = global_min_client_dates$anon_ID,
    enrollment_date = apply(global_min_client_dates[, 2:ncol(global_min_client_dates)], 1, min, na.rm = TRUE)
  ) %>% 
  mutate(enrollment_month = floor_date(ymd(enrollment_date), "month"))

monthly_util <- 
  monthly_util %>%
  left_join(., global_min_client_dates) %>%
  mutate(active_period = ifelse(month >= enrollment_month, 1, 0)) %>%
  filter(active_period == 1)
```

Example output (some column left out for display purposes)
```{r}
monthly_util %>% 
  filter(anon_ID == "210") %>%
  select(anon_ID, month, starts_with("used"), num_svcs_used)

```

Below is an example of why the master calendar was helpful. Client 1 used very little services, and its important to have all the months where no services where used.

```{r}

monthly_util %>% 
  filter(anon_ID == 1) %>%
  select(anon_ID, month, starts_with("used"), num_svcs_used) %>%
  print(n = 40)

```

#### active clients

the function below is an attempt to define 'active clients'. Since adding up the various encounters poses problems since they are different scales, and units (minutes vs pounds vs visits) I'm thinking this can be used a proxy for utilization? it can indicate how many clients consistently use services over time.

The logic is, for each client:

 * calculate a 3 month rolling mean of services used
 * check if that rolling mean is at least 1, meaning on avg they use 1 type of service offered each month for the past 3 months

```{r}
get_active_clients <-
  function(data, client, lookback = 2){

    data %>%
      ungroup() %>%
      filter(anon_ID == {{client}}) %>%
      mutate(lookback_mean = slider::slide_dbl(num_svcs_used, mean, .before = lookback, .after = 0),
             active_client = ifelse(lookback_mean >= 1, 1, 0))
  }

client_ids <- 
  sort(unique(monthly_util$anon_ID))

active_clients <- 
  map_dfr(client_ids, ~get_active_clients(data = monthly_util, client = .x)) 
```

using this methodology, Client 210 was an active client each month: 

```{r}
active_clients %>% 
  filter(anon_ID == "210") %>%
  select(month, anon_ID, num_svcs_used, lookback_mean, active_client) %>%
  print(n = 40)
```

but Client 1 dropped out of active client status: 

```{r}
active_clients %>% 
  filter(anon_ID == "1")%>%
  select(month, anon_ID, num_svcs_used, lookback_mean, active_client) %>%
  print(n = 40)
```

Plotting the overall active clients each month shows the impact of COVID 19
```{r}

map_dfr(client_ids, ~get_active_clients(data = monthly_util, client = .x)) %>%
  ungroup() %>%
  group_by(month) %>%
  summarise(active_clients = sum(active_client)) %>%
  ggplot(., aes(x = month, y = active_clients)) + 
  geom_point() + 
  geom_line() + 
  labs(x = 'month', y = "Count", title = 'Active Clients per Month')

```   

plotting active clients per month, poverty status and minority status shows how active minority clients with poverty = yes showed almost not change through the pandemic 
```{r}

map_dfr(client_ids, ~get_active_clients(data = monthly_util, client = .x)) %>%
  ungroup() %>%
  group_by(month, poverty, minority) %>%
  summarise(active_clients = sum(active_client)) %>%
  ggplot(., aes(x = month, y = active_clients, color = poverty)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(vars(minority)) + 
  labs(x = 'month', y = "Count", title = 'Active Clients per Month')

```   

#### other ways to measure utilization

Another way to measure utilization is to add up each type of encounter per client per month. 

It appears that volunteer encounters for minority clients is not  recovering the same as non-minority clients after the shock of the pandemic.

```{r}
monthly_util %>%
  select(anon_ID, month, minority, ends_with("encounters")) %>% 
  as_tibble() %>% 
  pivot_longer(4:ncol(.)) %>%
  group_by(month, name, minority) %>%
  summarise(total_encounters = sum(value, na.rm = T),
            mean_encounters = mean(value, na.rm = T),
            sd_util = sum(value, na.rm = T)) %>%
  ggplot(aes(x = month, y = total_encounters, color = minority)) +
  geom_point() +
  geom_line() + 
  facet_wrap(vars(name), scales = 'free') + 
  labs(x = "month", y = "Total monthly encounters", title = " Encounters by service and minority status")

```

It does seem like ElderNet was able to successfully transition to remote contact when the pandemic hit:

```{r}

care_mgmt_detail <-
  care_mgmt %>%
  mutate(month = as.Date(floor_date(assistance_date, 'month'))) %>%
  group_by(anon_ID, assistance_category, month) %>%
  summarise(total_care_mgmt_encounters = n(),
            total_care_mgmt_mins = sum(amount, na.rm = T),
            mean_care_mgmt_mins = mean(amount, na.rm = T))

client_info %>%
  left_join(., care_mgmt_detail, by = c("anon_ID", "month")) %>%
  select(anon_ID, month, minority, assistance_category, ends_with("encounters")) %>% 
  mutate(across(where(is.numeric), ~(ifelse(is.na(.x), 0, .)))) %>%
  group_by(month, assistance_category) %>%
  summarise(total_encounters = sum(total_care_mgmt_encounters, na.rm = T),
            mean_encounters = mean(total_care_mgmt_encounters, na.rm = T),
            sd_util = sum(total_care_mgmt_encounters, na.rm = T)) %>%
  ggplot(aes(x = month, y = total_encounters, color = assistance_category)) +
  geom_point() +
  geom_line() + 
  # facet_wrap(vars(name), scales = 'free') + 
  labs(x = "month", y = "monthly encounters", title = "Care mgmt encounters by service")
```

