---
title: "Food Pantry Utilization"
author: "Katrina Gutierrez"
date: "2/23/2022"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Food Pantry Utilization

```{r echo = FALSE, message = FALSE }
# load data
pantry <- read.csv("./../../../data/pantry_anonymized.csv")
client_data <- read.csv("./../../../data/client_info_anonymized.csv")

library(tidyverse)
library(lubridate)
```

## Some Clients Have Multiple Entries Logged for the Same Date

There may be some duplicate records in the food pantry data.  However, it's possible this data is genuine because a client may have picked up multiple bags in one visit and it was documented as separate records.  

12 clients clients have duplicate records.  For reference, there are 414 clients with food pantry data. As an example, anon_ID 15 looked like it had two of each of its records.  

Regardless, it's possible the duplicate records don't really matter if we focus on the number of visits (distinct dates) and when they ocurred.

```{r echo = FALSE, message = FALSE, results = "hide"}
# records with duplicates
duplicates <- pantry[duplicated(pantry),] 

# there are 132 records that have at least one duplicate of itself
nrow(duplicates)
# if we were to drop the duplicates, we would drop 2.1% of the data
1-nrow(unique(pantry))/nrow(pantry)

# 12 clients have duplicate records
length(unique(duplicates$anon_ID))
# clients impacted:
unique(duplicates$anon_ID)
# for reference there are 414 unique clients with pantry data
length(unique(pantry$anon_ID))

#as an example of some duplicates for anon_ID 15
# it looks like all records were there twice 
filter(pantry, anon_ID == "15") %>% arrange(assistance_date) %>% head(10)

```

```{r echo = FALSE, message = FALSE, results = "hide"}
# change date to date type
pantry <- pantry %>%
  mutate(assistance_date = as.Date(assistance_date, "%m/%d/%Y %H:%M")) 
  
```

# How often do clients visit the food pantry?

18.6% of clients who visited the food pantry at all only visited once.  
42.27% of clients who visited the food pantry went at least 10 times.  

```{r }
# creating a table organized by client
pantry_by_client <- pantry %>%
  group_by(anon_ID) %>%
  summarise(
    num_visits = length(unique(assistance_date)), 
    first_visit = min(assistance_date),
    last_visit = max(assistance_date)
  ) 

# plot
hist(pantry_by_client$num_visits,
     xlab = "Number of Pantry Visits",
     main = "How many clients visited the pantry X number of times")

# % visited more than 10 times
nrow(pantry_by_client %>% filter(num_visits >= 10))/nrow(pantry_by_client)

# % visited only 1 time
nrow(pantry_by_client %>% filter(num_visits == 1))/nrow(pantry_by_client)

```

# How long did clients use the food pantry?

Histogram of total length of time clients used the food pantry (last visit - first visit)

Since our data is limited, it's possible that this is an underrepresentation of total duration.

```{r }
# getting total duration of pantry usage per client
pantry_by_client <- pantry_by_client %>%
  mutate(usage_length =  as.numeric(difftime(last_visit, first_visit, units = "days")))

# plot
hist(pantry_by_client$usage_length,
     main = "How long did clients use the food pantry",
     xlab = "Usage Length (in days)")

```


## Time in Between Pantry visits

For clients who visited more than once, most of them visited the pantry within two months of their last visit on average.

I found the length of time since the previous visit for each visit.  Then, grouping by client, I found the average time between visits.

```{r }
# creates a table to get avg time b/w visits
avg_time_between_visits <- unique(pantry) %>% # remove duplicate rows 
  group_by(anon_ID)%>%  
  arrange(anon_ID, assistance_date) %>%
  filter(n() > 1) %>%
  mutate(Difference = difftime(assistance_date, lag(assistance_date), units = "days") )  %>%
  summarize(mean_time = mean(Difference, na.rm=TRUE)) 

# plot
boxplot(as.numeric(avg_time_between_visits$mean_time), 
        log = "y",
        main = "Distribution of Average Time Between Pantry Visits",
        ylab = "Avg Time Between Pantry Visits",
        sub = "For clients who had more than 1 visit (N = 337)")

```

# Poverty data

68.36% of clients who used the food pantry were in poverty

```{r }
length(unique(pantry$anon_ID)) # unique food pantry clients

pantry <- merge(pantry, client_data, by = "anon_ID", all.x = TRUE)

pantry_in_poverty = filter(pantry, poverty == "Yes")

length(unique(pantry_in_poverty$anon_ID)) # 283 food pantry clients in poverty

283/414



```


