---
title: Using the `tidycensus` R Package to Analyze the Population Demographics of
  Montgomery County, PA
author: "Kathrine McAulay"
date: "3/27/2022"
output: github_document
---

```{r setup}
library(tidycensus)
library(forcats)
# To recreate these anlyses you will need your own API key
# https://api.census.gov/data/key_signup.html
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

### Resources
The code in this document is based on examples found here:

* https://walker-data.com/census-r/index.html
* 

## Montgomery County, PA

ElderNet serves the Lower Merion and Narberth Townships of Montgomery County, PA. We can use the `tidycensus` package to look at the population demographics of this region based on the US Census Bureau data.

One way census reported variable we can look at is "Median household income in the past 12 months (in 2020 inflation-adjusted dollars)", represented by variable "B19013_001".

Setting `geography` to "county subdivision" we can see subtotals for each Township whin the County, including Lower Merion and Narberth.

```{r}
montgomery_acs_2020 <- get_acs(
  geography = "county subdivision",
  state = "PA",
  county = "Montgomery",
  variables = "B19013_001",
  geometry = TRUE,
  year = 2020
  )
```

We can then use `geom_sf` to compare median household income across each Township in the County:

```{r}
montgomery_income_map <- ggplot(montgomery_acs_2020,
                                aes(fill = estimate, color = estimate)) +
  geom_sf() +
  scale_fill_viridis(direction = -1, option = "I") + 
  scale_color_viridis(direction = -1, option = "I") +
  labs(title = "Median Income Across Montgomery County, PA",
       subtitle = "Data source: US Census Bureau, 2016-2020 ACS",
       fill = "Estimated Median\nIncome (USD)") +
  guides(color = "none") +
  theme_void() +
  theme(plot.title = element_text(size = 17),
        plot.subtitle = element_text(size = 13))
```

! [Median Income Montgomery County] (analyses/team1/kathrine_m/images/montgomery_income_map.png)


### Ra
